Magic-free Code Extraction of Generic Programs and Recursion Schemes in Coq

### 2024-03-18 paper/paper.tex
intro

Not many approaches mechanise structured recursion
schemes in proof assistants.  Recently, Abreu et al.~\cite{AbreuDHJMS23} encode
an algebraic approach to divide-and-conquer computations, in which termination
is entirely enforced by the typing discipline of the recursion scheme. However,
although their approach solves the problem of termination proofs, as well as
the performance of the code that is run \emph{within Coq}, it no longer allows
the extraction of idiomatic OCaml code.  By this, we refer to code
that (1) does not preserve the recursive structure of common implementations;
and (2) lead to unsafe casts \ocaml{Obj.magic} in the generated code. 
In OCaml, \ocaml{Obj.magic} allows for unsafe casts between types.
Although the use of \ocaml{Obj.magic} should be safe in the code
extracted by Coq, it has been proven that, for higher-order programs, simple
interoperations can lead to incorrect behaviour or even
segfaults~\cite{forster:hal-04329663}. Therefore, the avoidance of axioms and
\ocaml{Obj.magic} means that the extracted code will have stronger
safety guarantees, and will allow the safe interoperation with other OCaml
code.

@@

Within the domain of program optimization, program calculation serves as a
well-established programming technique aimed at deriving efficient programs
from their naive counterparts through systematic program
transformation~\cite{gibbonsSquiggol}. This area has been extensively explored
over the years. Tesson et al.\ demonstrated the efficacy of leveraging Coq to
establish an approach for implementing a robust system dedicated to verifying
the correctness of program transformations for functions that manipulate
lists~\cite{10.1007/978-3-642-17796-5_10}. Murata and Emoto went further and
formalised recursion schemes in Coq~\cite{MurataE19}. Their development does
not include hylomorphisms and dynamorphisms, and relies on the functional
extensionality axiom, as well as further extensionality axioms for each
coinductive datatype that they use. They do not discuss the extracted OCaml
code from their formalisation.  Larchey-Wendling and Monin encode recursion
schemes in Coq, by formalising computational graphs of
algorithms~\cite{larchey2022braga}. Their work does not focus on encoding
generic recursion schemes, and proving their algebraic laws.  Castro et
al.~\cite{farmsCastro} encode the laws of hylomorphisms as part of the type
system of a functional language to calculate parallel programs from
specifications. Their work focuses on parallelism, and they do not formalise
their approach in a proof assistant, and the laws of hylomorphisms are axioms
in their system.

>>>>

Not many approaches mechanise structured recursion
schemes in proof assistants.  Recently, Abreu et al.~\cite{AbreuDHJMS23} encode
an algebraic approach to divide-and-conquer computations, in which termination
is entirely enforced by the typing discipline of the recursion scheme. However,
although their approach solves the problem of termination proofs, as well as
the performance of the code that is run \emph{within Coq}, it no longer allows
the extraction of idiomatic OCaml code.  By this, we refer to code
that (1) does not preserve the recursive structure of common implementations;
and (2) lead to unsafe casts \ocaml{Obj.magic} in the generated code. 
In OCaml, \ocaml{Obj.magic} allows for unsafe casts between types.
Although the use of \ocaml{Obj.magic} should be safe in the code
extracted by Coq, it has been proven that, for higher-order programs, simple
interoperations can lead to incorrect behaviour or even
segfaults~\cite{forster:hal-04329663}. Therefore, the avoidance of axioms and
\ocaml{Obj.magic} means that the extracted code will have stronger
safety guarantees, and will allow the safe interoperation with other OCaml
code. The absence of \ocaml{Obj.magic} provides even stronger guarantees to the
extracted code. In particular, since the code that we extract from Coq is
implemented in a total language, we know that there can be no $\bot$ in its
definition. This allows the programmer to use ``fast-and-loose'' reasoning on
the extracted code~\cite{DanielssonHJG06}, which would otherwise be unsafe if
the code contained \ocaml{Obj.magic}. For example, the programmer would need to
manually check that the inputs satisfy the necessary assumptions which would be
impossible to enforce in the presence of unsafe casts. Overall, the absence of
\ocaml{Obj.magic} makes it easer to reason about, and interoperate with
extracted code.

@@

Within the domain of program optimization, program calculation serves as a
well-established programming technique aimed at deriving efficient programs
from their naive counterparts through systematic program
transformation~\cite{gibbonsSquiggol}. This area has been extensively explored
over the years. Tesson et al.\ demonstrated the efficacy of leveraging Coq to
establish an approach for implementing a robust system dedicated to verifying
the correctness of program transformations for functions that manipulate
lists~\cite{10.1007/978-3-642-17796-5_10}. Murata and Emoto went further and
formalised recursion schemes in Coq~\cite{MurataE19}. Their development does
not include hylomorphisms and dynamorphisms, and relies on the functional
extensionality axiom, as well as further extensionality axioms for each
coinductive datatype that they use. They do not discuss the extracted OCaml
code from their formalisation.  Larchey-Wendling and Monin encode recursion
schemes in Coq, by formalising computational graphs of
algorithms~\cite{larchey2022braga}. Their work does not focus on encoding
generic recursion schemes, and proving their algebraic laws.  Castro-Perez et
al.~\cite{farmsCastro} encode the laws of hylomorphisms as part of the type
system of a functional language to calculate parallel programs from
specifications. Their work focuses on parallelism, and they do not formalise
their approach in a proof assistant, and the laws of hylomorphisms are axioms
in their system.
